---
title: "R Commands for Causal Inference"
author: "Robin Evans"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{R Commands for Causal Inference}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


# R Commands

This document contains some example code blocks, that can be used with the
exercise on the `bcrot` dataset.

```{r setup}
library(APTSCausalInference)
library(ggplot2)
data(bcrot)
```

## Plotting

The below gives histograms comparing quality of life under hormone treatment 
and control.

```{r plotqol, eval = FALSE}
plot.qol <- ggplot(bcrot, aes(x=qol, after_stat(density)), fill=as.factor(hormon)) +
    geom_histogram(aes(fill=as.factor(hormon)), color=c("#e9ecef"), binwidth = 2) +
    facet_grid(as.factor(hormon) ~ .) +
    labs(x = "Quality of life", y = "Density") +
    scale_fill_manual(values=c("#69b3a2", "#337CA0"),
                      name="Hormonal\ntreatment",
                      breaks=c("0", "1"),
                      labels=c("no", "yes")) +
    theme_light()
plot.qol
```
This gives a density plot for the proposensity score (`ps`).

```{r density, eval=FALSE}
# plot density
ggplot(bcrot, aes(ps, fill = as.factor(hormon))) +
  geom_density(alpha = 0.5) +
  labs(x = "Propensity score",
       y = "Density",
       fill = "Hormone\ntreatment") +
  scale_fill_manual(values=c("#69b3a2", "#337CA0"),
                    name="Hormonal\ntreatment",
                    breaks=c("0", "1"),
                    labels=c("no", "yes")) +
  theme_light()
```

## Love Plots

Love plots can be obtained with the command `cobalt::love.plot`.  Make sure you 
put in the right entry for the right-hand side of the formula, and for the 
`data` argument!

```{r love_plot, eval=FALSE}
cobalt::love.plot(hormon_i ~ XXX , 
                  data =  XXX  , 
                  weights = bcrot$w,
                  s.d.denom = "pooled",
                  thresholds = c(m = .1))
```

## Subsetting and Filtering

```{r dplyr, eval=FALSE}
# restrict dataset and transformation
bcrot2 <- bcrot %>% filter(age >= 40,      # keep only individuals over 40
                           XXX) %>%   # XXXX FILL IN! w/>= 1 cancerous node
                    mutate(age2 = age * age) %>%    # add age²
                    select(-c(ps, w))      # drop propensity score and weights
```


## Regression Standardization

```{r stdReg, eval=FALSE}
fit <- glm(qol ~ hormon*age + XXXX,  ## fill in other variables!
           data = bcrot2)
fit.std <- stdGlm(fit = fit, data = as.data.frame(bcrot2), X = "hormon")
print(summary(fit.std))
plot(fit.std)

# Confidence interval: Mean difference 
summary(fit.std, contrast = "difference", reference = "0")
```

## AIPW

```{r AIPW, eval=FALSE}
# Build SuperLearner libraries for outcome (Q) and exposure models (g)
sl.lib <- c("SL.mean","SL.glm")  

# Construct an aipw object for later estimations 
AIPW_SL <- AIPW::AIPW$new(Y = bcrot2$qol,
                          A = bcrot2$hormon_i,
                          W = subset(bcrot2, select = c(XXX)),   # fill in!
                          Q.SL.library = sl.lib,
                          g.SL.library = sl.lib,
                          k_split = 10,
                          verbose = TRUE)

# Fit the data to the AIPW object and check the balance of propensity scores
raipw <- AIPW_SL$fit()$summary()
AIPW_SL$fit()$plot.p_score()$plot.ip_weights()

# Truncate weights (default truncation is set to 0.025)
AIPW_SL$fit()$summary(g.bound = c(0.05, 0.95))$plot.p_score()$plot.ip_weights()

# Calculate average treatment effects among the treated/untreated (controls) (ATT/ATC)
AIPW_SL$stratified_fit()$summary()


# extract weights for loveplots
# AIPW_SL$plot.ip_weights()
bcrot2$aipw <- AIPW_SL$ip_weights.plot$data$ip_weights
```