---
title: "Causal Inference"
author: "V Didelez, R Foraita"
date: "19 08 2025"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Causal Inference}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  eval = TRUE,
  collapse = TRUE,
  fig.width = 6)
```

## Load packages
```{r, warning=FALSE, message=FALSE}
library(APTSCausalInference)
# required for data management and plots
library(data.table)
library(dplyr)
library(ggplot2)

# required for analysis
library(AIPW)
library(cobalt)
library(ipw)
library(sandwich)
library(stdReg)
library(SuperLearner)
library(survey)
```
## Load the Rotterdam Breast Cancer data set
```{r load data}
# data(bcrot)
load("../data/bcrot.RData")
```

## Descriptive Analysis
Compare descriptively QOL for those who do and do not take hormonal therapy 

```{r descriptives}
# qol ~ hormon
plot.qol <- ggplot(bcrot, aes(x=qol, after_stat(density)), fill=as.factor(hormon)) +
    geom_histogram(aes(fill=as.factor(hormon)), color=c("#e9ecef"), binwidth = 2) +
    facet_grid(as.factor(hormon) ~ .) +
    labs(x = "Quality of life", y = "Density") +
    scale_fill_manual(values=c("#69b3a2", "#337CA0"),
                      name="Hormonal\ntreatment",
                      breaks=c("0", "1"),
                      labels=c("no", "yes")) +
    theme_light()
plot.qol


# age ~ hormon
plot.age <- ggplot(bcrot, aes(x=age, after_stat(density)), fill=as.factor(hormon)) +
    geom_histogram(aes(fill=as.factor(hormon)), color=c("#e9ecef"), binwidth = 2) +
    facet_grid(as.factor(hormon) ~ .) +
    labs(x = "Age", y = "Density") +
    scale_fill_manual(values=c("#69b3a2", "#337CA0"),
                      name="Hormonal\ntreatment",
                      breaks=c("0", "1"),
                      labels=c("no", "yes")) +
    theme_light()
plot.age


# Lymph nodes ~ hormon
plot.nodes <- ggplot(bcrot, aes(x=nodes, after_stat(density)), fill=as.factor(hormon)) +
    geom_histogram(aes(fill=as.factor(hormon)), color=c("#e9ecef"), binwidth = 1) +
    facet_grid(as.factor(hormon) ~ .) +
    labs(x = "Number of positive lymph nodes", y = "Density") +
    scale_fill_manual(values=c("#69b3a2", "#337CA0"),
                      name="Hormonal\ntreatment",
                      breaks=c("0", "1"),
                      labels=c("no", "yes")) +
    theme_light()
plot.nodes

ggplot(bcrot, aes(x=enodes, after_stat(density)), fill=as.factor(hormon)) +
  geom_histogram(aes(fill=as.factor(hormon)), color=c("#e9ecef"), binwidth = 0.1) +
  facet_grid(as.factor(hormon) ~ .) +
  labs(x = "Number of positive lymph nodes (transformed: exp(-0.12 * nodes))", 
       y = "Density") +
  scale_fill_manual(values=c("#69b3a2", "#337CA0"),
                    name="Hormonal\ntreatment",
                    breaks=c("0", "1"),
                    labels=c("no", "yes")) +
  theme_light()


# PgR (fmol/l), log ~ hormon
ggplot(bcrot, aes(x=pr_1, after_stat(density)), fill=as.factor(hormon)) +
  geom_histogram(aes(fill=as.factor(hormon)), color=c("#e9ecef"), binwidth = 0.5) +
  facet_grid(as.factor(hormon) ~ .) +
  labs(x = "PgR [fmol/l] (transformed: log(pr)", y = "Density") +
  scale_fill_manual(values=c("#69b3a2", "#337CA0"),
                    name="Hormonal\ntreatment",
                    breaks=c("0", "1"),
                    labels=c("no", "yes")) +
  theme_light()

```

## Analysis
### Linear Model
We compare an unadjusted naive linear model (2-groups) with a naive adjusted main-effects-only model (this is too simplistic to be plausible; and it is not robust). We can see that the unadjusted analysis is (likely) very biased - the effect reverses upon adjustment. Note that the analysis does not alert us to any positivity issues as it simply extrapolates.
```{r linear model unadjusted}
# Unadjusted linear regression
lm.ua <- lm(qol ~ hormon, data = bcrot)
summary(lm.ua)

```

```{r linear model adjusted}
# Main-effects linear regression
lm.a <- lm(qol ~ hormon + age + enodes + pr_1, data = bcrot)
summary(lm.a)

```
### Propensity score (PS)
We saw that positivity is an issue with nodes=0 and younger ages. Here we look at how to detect this from the propensity score (PS). We begin with a simple main-effects-only PS model.
```{r, echo = TRUE}
ps <- glm(hormon ~ age + enodes + pr_1, 
          data = bcrot, 
          family = binomial(link="logit"))

# add fitted values to data set
bcrot$ps <- fitted(ps)

# plot density
ggplot(bcrot, aes(ps, fill = as.factor(hormon))) +
  geom_density(alpha = 0.5) +
  labs(x = "Propensity score",
       y = "Density",
       fill = "Hormone\ntreatment") +
  scale_fill_manual(values=c("#69b3a2", "#337CA0"),
                    name="Hormonal\ntreatment",
                    breaks=c("0", "1"),
                    labels=c("no", "yes")) +
  theme_light()
 
 
# Histogram
ggplot(data = bcrot, aes(x = ps, after_stat(density), fill = as.factor(hormon))) +
  #geom_histogram(alpha = 0.5, position = "identity", binwidth = 0.05) +
  geom_histogram(aes(fill=as.factor(hormon)), color=c("#e9ecef"), binwidth = 0.05) +
  facet_grid(hormon ~ .) +
  labs(x = "Propensity score", y = "Density") +
  scale_fill_manual(values=c("#69b3a2", "#337CA0"),
                    name="Hormonal\ntreatment",
                    breaks=c("0", "1"),
                    labels=c("no", "yes")) +
  theme_light()
```

### Inverse Probability Treatment Weighting 
We also find that we have some extreme weights (though, I have seen worse). The weighted data does not achieve balance in age and enodes; this could be due to lack of positivity and/or because the PS model is bad.
```{r IPTW}
bcrot$w <- (as.numeric(as.character(bcrot$hormon)) / bcrot$ps) + 
           ((1 - as.numeric(as.character(bcrot$hormon))) / (1 - bcrot$ps))
summary(bcrot$w)

# Plot inverse probability weights vs. index
ggplot(bcrot, aes(x = 1:nrow(bcrot), y = w)) + 
  geom_point()  +
  xlab(" ") +
  ylab(" ") + 
  ylim(0, 65) +
  theme_minimal()


# Covariate balance plot (love plot - named after Thomas E. Love)
cobalt::love.plot(hormon ~ age + enodes + pr_1, 
                  data = bcrot, 
                  weights = bcrot$w,
                  s.d.denom = "pooled",
                  thresholds = c(m = .1))

```


### Use subset
We decide that the effect is ill-defined for nodes=0 and age<40; thus we create a more meaningful subset of the population for which to compare treatment with no-treatment. Note that we lose more than half (ca. 1500) of the untreated, but only 7 of the treated patients.
```{r subset}
plot.age
plot.nodes

# restrict dataset and transformation
bcrot2 <- bcrot %>% filter(age >= 40,
                           nodes > 0) %>%
                    mutate(age.2 = age * age) %>%    # add age²
                    select(-c(ps, w))  

table(bcrot2$hormon)
table(bcrot$hormon)


# Main-effects and adjustes linear regression models using the subset
lm.ua2 <- lm(qol ~ hormon, data = bcrot2)
summary(lm.ua2)
lm.a2 <- lm(qol ~ hormon + age + enodes + pr_1, data = bcrot2)
summary(lm.a2)
```


### Estimate Marginal structural model "by hand"
First fit new more complex PS model (as we did not achieve balance earlier) and check balance. Next fit MSM by weighted lm; note the underestimated standard errors! Obtain valid SEs by sandwich estimation.

```{r MSM by hand}
ps <- glm(hormon ~ age + age.2 + enodes + age*pr_1,
          data = bcrot2, 
          family = binomial(link="logit"))
bcrot2$ps <- fitted(ps)
bcrot2$w <- (as.numeric(as.character(bcrot2$hormon)) / bcrot2$ps) + 
           ((1 - as.numeric(as.character(bcrot2$hormon))) / (1 - bcrot2$ps))

# Checking balance
cobalt::love.plot(hormon ~ age + age.2 + enodes + age*pr_1, 
                  data = bcrot2, 
                  weights = bcrot2$w,
                  s.d.denom = "pooled",
                  thresholds = c(m = .1))

cobalt::bal.plot(ps, "pr_1", which = "both", colors = c("#69b3a2", "#337CA0"), data=bcrot2)
cobalt::bal.plot(ps, "age", which = "both", colors = c("#69b3a2", "#337CA0"), data=bcrot2)
cobalt::bal.plot(ps, "enodes", which = "both", colors = c("#69b3a2", "#337CA0"), data=bcrot2)


# Weights based on ~age + age.2 + enodes + age*pr_1
model_w <- lm(qol ~ hormon , weights = w, data = bcrot2)
summary(model_w)

# Variance estimation using the robust sandwich variance estimator
(sandwich_se <- diag(sandwich::vcovHC(model_w, type = "HC"))^0.5)
# confidence interval
sandwichCI <- c(coef(model_w)[2] - 1.96 * sandwich_se[2],
                      coef(model_w)[2] + 1.96 * sandwich_se[2])
```


### Estimate Marginal structural model using the IPW package
With package `ipw` can do the whole MSM in one. Also has an inbuilt plot for checking weights distribution.

```{r MSM IPW}
ipw2 <- ipw::ipwpoint(exposure = hormon, 
                      family = "binomial", link = "logit",
                      denominator = ~ age + age.2 + enodes + age*pr_1,
                      data = bcrot2)
bcrot2$ipw <- ipw2$ipw.weights

# Plot Inverse Probability Weights
summary(ipw2$ipw.weights)
ipw::ipwplot(weights = ipw2$ipw.weights, 
             logscale = FALSE,
             main = "Stabilized weights", 
             xlab = "Weights",
             xlim = c(1, 10))

# Marginal structural model for the causal effect of hormon on qol
# corrected for confounding using inverse probability weighting
# with robust standard error from the survey package.
model_sm2 <- survey::svyglm(qol ~ hormon,
                            design = svydesign(~ 1, 
                                               weights = ~ipw,
                                               data = bcrot2))
summary(model_sm2)
confint(model_sm2)
```


### Compare confidence intervals
```{r ci}
msm.out <- rbind(cbind(coef(model_w), confint(model_w))[2,],
                 c(coef(model_w)[2], sandwichCI),
                 cbind(coef(model_sm2), confint(model_sm2))[2,])
dimnames(msm.out) <- list(c("MSM, naive", "MSM, robust SE (sandwich):","MSM, robust SE (ipw):"), 
                          c("est", colnames(msm.out)[2:3]))
msm.out
```



### Investigate (extreme) weights - truncation?
```{r extreme weights}
# truncate weights
wq <- quantile(bcrot2$ipw, probs = c(0.025, 0.975)) 
bcrot2$wt <- if_else(bcrot2$ipw < wq[1], wq[1], bcrot2$ipw)
bcrot2$wt <- if_else(bcrot2$wt  > wq[2], wq[2], bcrot2$wt)

summary(bcrot2$ipw)
summary(bcrot2$wt)

# Plot weight vs. index without and with truncation
ggplot(bcrot2, aes(x = 1:nrow(bcrot2), y = ipw)) + 
  geom_point()  +
  xlab(" ") +
  ylab(" ") + 
  ylim(0, 65) +
  labs(title = "Not truncated") +
  theme_minimal()

ggplot(bcrot2, aes(x = 1:nrow(bcrot2), y = wt)) + 
  geom_point()  +
  xlab(" ") +
  ylab(" ") + 
  ylim(0, 10) +
  labs(title="Truncated weights") + 
  theme_minimal()


# loveplot without and with truncation
# truncation loses some balance - no real need for truncation here
cobalt::love.plot(hormon ~ age + age.2 + enodes + age*pr_1, 
                     data = bcrot2, 
                     weights = bcrot2$ipw,
                     s.d.denom = "pooled",
                     thresholds = c(m = .1),
                     title = "Not truncated"
                     )

cobalt::love.plot(hormon ~ age + age.2 + enodes + age*pr_1, 
                      data = bcrot2, 
                      weights = bcrot2$wt,
                      s.d.denom = "pooled",
                      thresholds = c(m = .1),
                      title = "Truncated weights")
```



### Regression standardization
Using the correct outcoem model (known because we generated QoL from a known model), we estimate the population average effect (and its standard errors) by standardization with package stdReg.
```{r regression standardization}
fit <- glm(qol ~ hormon*age + hormon*age.2 + hormon*enodes + hormon*pr_1 + age*pr_1,
           data = bcrot2)
fit.std <- stdGlm(fit = fit, data = as.data.frame(bcrot2), X = "hormon")
print(summary(fit.std))
plot(fit.std)

# Confidence interval: Mean difference 
summary(fit.std, contrast = "difference", reference = "0")
```



### Double machine learning - Augmented Inverse Probability Weighting
Here we use AIPW with some default settings. 
```{r DML, warning=FALSE}
# Build SuperLearner libraries for outcome (Q) and exposure models (g)
sl.lib <- c("SL.mean","SL.glm")  

# Construct an aipw object for later estimations 
AIPW_SL <- AIPW::AIPW$new(Y = bcrot2$qol,
                          A = as.integer(as.character(bcrot2$hormon)),
                          W = subset(bcrot2, select = c("enodes", "age", "pr_1")), 
                          Q.SL.library = sl.lib,
                          g.SL.library = sl.lib,
                          k_split = 10,
                          verbose = TRUE)

# Fit the data to the AIPW object and check the balance of propensity scores
raipw <- AIPW_SL$fit()$summary()
AIPW_SL$fit()$plot.p_score()$plot.ip_weights()

# Truncate weights (default truncation is set to 0.025)
AIPW_SL$fit()$summary(g.bound = c(0.05, 0.95))$plot.p_score()$plot.ip_weights()

# Calculate average treatment effects among the treated/untreated (controls) (ATT/ATC)
AIPW_SL$stratified_fit()$summary()


# extract weights for loveplots
# AIPW_SL$plot.ip_weights()
bcrot2$aipw <- AIPW_SL$ip_weights.plot$data$ip_weights

# loveplots + AIPW
cobalt::love.plot(hormon ~ age*enodes*pr_1 + 
                           age.2*enodes*pr_1, 
                  data = bcrot2, 
                  weights = bcrot2$aipw,
                  s.d.denom = "pooled",
                  thresholds = c(m = .1))
```



### Compare results in the restricted sample
This is a summary of all the different methods - of these we know that regression standardization is correctly specified; we do not know this for any of the other methods; but we know the true effect for our synthetic QoL outcome. Note the much larger uncertainty for IPW.

```{r comparison}
out <- rbind(c(2.07, rep(NA,3)), 
             c(coef(summary(lm.a2))[2,1:2], confint(lm.a2)[2,]),
             c(summary(model_sm2)$coefficients[2,1:2], confint(model_sm2)[2,]),
             summary(fit.std, contrast = "difference", reference = "0")$est.table[2,],
             raipw$estimates$RD)
dimnames(out) <- list(c("True ATE", "LM", "IPW", "RS", "AIPW"),
                      c("Estimate", "SE", "lower", "upper"))
print(round(out,3))
```